Disclosing funding sources for open access publication fees: the Open APC

### 2016-06-05 article.Rmd
conclusion first draft

The analysis of APC spending is difficult. The total spending is fragmented across the budgets of grant agencies, research institutions, and libraries, or is covered by personal budgets. Asking 9,645 authors from various disciplines how they financed publication fees, a comprehensive survey in 2010 revealed that the majority of the respondents had access to research funding or institutional support to cover these charges. By contrast, 12 % paid publication fees individually [@soap]. These findings are consistent with that of other studies, adding that sources of funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets, on the other hand, are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

@@

Hybrid journals substantially add to this complexity of open access funding [@Kingsley_2014; @Pinfield_2015]. These journals, allowing articles to be published immediately as open access after a charge was paid, rely both on subscriptions and publication fees as revenue sources. Although the uptake of open access through hybrid journals was described as lower and more expensive compared to that of fully open access journals [@Solomon_2012; @Bj_rk_learned_2014], this model has gained attention through recent science policies, notably because of open access policies from the UK [@Pinfield_Review_2015].

To address these problems of fragmented spending on publication fees and of in-transparency about what was being paid, some European research funders and research performing institutions have recently begun to disclose their payment for publication as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc_12] and the Austrian Science Fund FWF [@fwf_apc_13]. The not-for-profit company Jisc followed by collecting data from UK universities [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report what was being spent along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates and persistent identifiers to the full text including the Digital Object Identifier (DOI) [@Neylon_2014; @woodward_2014]. Parts of Jisc's cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variances, as well as a varying number of articles UK universities supported between 2007 - 2014, confirming earlier studies that collected price information from journal websites.

<!-- This complex situation of fee-based open access publishing creates difficulties for researchers and practitioners alike. Some European initiatives have therefore begun to collect and disclose their expenditures on open access journal articles as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@fwf_apc_13] and the Austrian Science Fund FWF [@wellcome_apc_12]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report payments made along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version, and a link to a deposit in a subject repository. Curatorial efforts 
focused on the disambiguation of publisher and journal titles as well as on detecting duplicates. In the case of the Wellcome Trust, crowd-sourcing data cleaning activities through a Google spreadsheet in combination with checks against bibliographic sources massively improved the spending data (see comments in @wellcome_apc).-->

@@

This paper focuses on how much German universities and research organisations spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for these charges. Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" program in 2011, only a few central funds existed [@eppelin_2012]. This is similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012]. The DFG has enforced a set of criteria grantees have to comply with  [@Fournier_2013], leading to similar implementations for open access supporting across German universities. These criteria exclude sponsoring of articles in hybrid journals and the funding of articles whose publication fee exceeds 2,000 €[^3]. Grantees agree not only to pay for APCs but also to find ways to improve the handling of those financial transactions. This includes central invoicing schemes and memberships, among others, agreed between university libraries and publishers that, in many cases, lead to discounts [@Fournier_2013].

Research institutes outside of universities, e.g. those institutes organised in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft, are not eligible for this DFG funding program. But in response, some organizations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The evolving institutional support structures to cover open access publication fees has led to calls for an unified approach towards supporting open access journal publishing in Germany. The Allianz der Wissenschaftsorganisationen[^10c], a science policy board representing all major research organisations in Germany, marked price transparency as one way to sustain an "adequate open access publication system" [@allianz]. Reflecting Austrian and UK initiatives to share institutional spending on open access publication fees as open data, as well as professional discussions on open access publishing, Bielefeld University Library began to openly share its payment of publication fees in May 2014. After engaging with the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11], other German institutions joined under the umbrella of the Open APC initiative soon after.

@@

The aim of the study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Using self-reported cost data from the Open APC initiative, the analysis focused on the amount that was being spent on publication fees, and compared these expenditures with data from related Austrian and UK initiatives, both in terms of size and the proportion of articles being published in fully and hybrid open access journals. We also asked how thoroughly self-reported articles were indexed in Crossref, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.

@@

We analysed self-reported cost data released by the Open APC initiative on May 13, 2016[^24] to assess institutional spending on open access publication fees in Germany. In addition to administrative data about the amount paid per article including value added tax, the reporting institution, and the year of invoicing, we used information about whether an article was published in a fully open access journal or in a hybrid journal as well as the DOI reported in the data-set.

We fetched bibliographic metadata for each article from Crossref on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from Crossref as well, this information was retrieved at the time when the participating institutions submitted their expenses. The Open APC initiative kept track of the date when these data-sets were submitted with Git, a version control system, increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub to be transparent over time. However, during these data collection activities Crossref had regularly updated metadata to represent ongoing mergers of publishing houses. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve updated metadata from Crossref for the whole Open APC data-set instead of re-using the historical publisher and journal information contained in the Open APC data-set.

@@

Data collection also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by Germany universities and research organisations with that of other initiatives, we consulted the openly available data-sets from the the Austrian Science Fund FWF [@fwf_apc_14; @fwf_apc_15], Jisc [@jisc_14; @jisc_15] and the Wellcome Trust [@wellcome_apc_13_14; @wellcome_apc_14_15]. For analysis, we obtained the overall publication fee spending on both fully and hybrid open access journal articles. In the case of FWF, we gathered the cost information from the accompanying spending reports. We used the spreadsheet data to calculate Wellcome Trust's and Jisc's spending, and converted the prices from GBP to € in accordance with the average foreign exchange reference rates provided by the European Central Bank. Our comparison of the open data initiatives focussed on the last two years 2014 and 2015. Because Wellcome Trust's spending was reported for the financial periods 2013 - 2014 and 2014 - 2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded articles from the analysis for which either cost information or journal type was missing. In the case of Jisc's 2014 data [@jisc_14], for instance, we excluded spending on 2,812 publications that amounted to 4.861.772 € from the analysis because no publication type was given in the data-set.

The data collection methods of the Open APC initiative and those of the other initiatives differed in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institutions were required to report, in the Wellcome Trust data a publication identifier was added as part of the automated compliance checks. Our first screening of the data-sets revealed that some articles lacked a DOI. For this reason, and as our main focus was institutional funding for publication fees in Germany, we decided to only compare German APC spending with that reported by other initiatives. We did not compare the distribution of spending over publishers and journal titles, or the indexing coverage in Crossref.

@@

After excluding payments being made for non-journal articles as well as articles invoiced in 2016, we retrieved `r nrow(my_apc)` open access journal articles `r length(unique(my_apc$institution))` German universities and research institutions financially supported between 2005 and 2015. As illustrated in Figure 1, payments made for open access journal articles increased over the years. While one institution supported `r filter(my_apc, period == min(my_apc$period)) %>% nrow()` articles in `r as.character(min(my_apc$period))`, the majority shared their expenditure from 2013 onwards. With `r max(apc_time$articles)` articles, the year `r apc_time[apc_time$articles == max(apc_time$articles),"period"]` was best represented in our data-set. However, the data-set contained `r my_apc %>% filter(period == 2015) %>% group_by(institution) %>% summarise(n()) %>% nrow()` institutions that contributed their cost data for 2015 at the time of this analysis, suggesting a time lag between payments made and reporting these spending to the Open APC initiative.

@@

Figure 3 presents the large price variations. Publication fees that were paid by the German universities and research organisations ranged between `r min(my_apc$euro)` € and `r max(my_apc$euro)` €  (SD `r sd(my_apc$euro)`). However, the average price paid varied somewhat during the period 2011 and 2015 (`r my_apc %>% filter(period == 2011) %>% summarise(mean(euro)) %>% round()` - `r my_apc %>% filter(period == 2015) %>% summarise(mean(euro)) %>% round()` €). Average publication fee spending per article in Germany was `r mean(my_apc$euro)` (Median = `r median(my_apc$euro)` €, SD = `r sd(my_apc$euro)` €).

@@

The number of APC payments per institutions varied considerably (see Table 1). With `r filter(my_apc, institution == "MPG") %>% nrow()` reported articles, the Max Planck Society contributed `r (filter(my_apc, institution == "MPG") %>% nrow() / nrow(my_apc)) * 100` % of the overall submissions. By contrast, several universities that only recently begun to set up support structures for fee-based open access journal articles reported low numbers of payments.

@@

Table 2 compares Open APC spending data with that of the Austrian FWF, as well as with Jisc's and Wellcome Trust's expenditure. Prices were converted according to the average Euro exchange rate of the examined periods, and gathered for both fully and hybrid open access journals. The comparison reveals that the Open APC initiative lacked cost information about hybrid journals, whereas the related Austrian and UK open data initiatives could report a large share of spending on these journals between 2014 and 2015. This situation presumably reflected the importance of the DFG as funding source for publication fees in Germany. The DFG funding policy excluded support for publications in hybrid journals. Over the years 2005 – 2015, `r my_apc %>% group_by(institution) %>% filter(is_hybrid == TRUE) %>% summarise() %>% nrow()` out of `r length(unique(my_apc$institution))` German universities and research institutions reported `r my_apc %>% filter(is_hybrid == TRUE) %>% nrow()` hybrid journal articles to the Open APC initiative, representing `r (my_apc %>% filter(is_hybrid == TRUE) %>% nrow() / my_apc %>% nrow()) * 100` % of the overall supported article volume. In terms of the number of supported articles and the amount being spent on publication fees, by contrast, the Open APC data-set provided the most comprehensive price information for fully open access journals compared to what the Austrian and UK initiatives had reported.

@@

Comparison of average prices suggests that publishing in hybrid journal was more expensive than in fully open access journals. Price differences between these two categories were also reported earlier, indicating that prices for fully open access journals were typically lower [@Solomon_2012; @Pinfield_2015]. In 2014 and 2015, the mean price for fully open access journals calculated from all data-sets was below the DFG price cap of 2,000 €.

@@

We used the DOI to automatically fetch publisher and journal names for each article from the Crossref REST API. Table 4 shows the top ten publishers in terms of payments made that represented `r apc_publisher %>% filter(!publisher == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the overall spending on publication fees. In total, payments were made to `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. In comparison with data from the UK, some open access publishers have a greater share on total spending. @Pinfield_2015, for instance, reported remarkably lower numbers for the open access publishers MPDI, Copernicus, and Hindawi.

@@

Most of publication fee spending in Germany was on articles published in Springer Nature journals, most likely profiting from the merge with the open access publisher BioMed Central in 2008, and that between the well-established publishers Springer Science + Business Media and Nature Publishing Group in 2015. Using the Crossref-Member-ID instead of the publisher name, we were able to differentiate between journals formely published by Springer Science + Business Media and Nature Publishing Group. In terms of articles, the majority of payments made were for publications in journals formerly associated with Springer Science + Business Media. Springer Science + Business Media journals accounted for `r nrow(springer)` articles, representing `r nrow(springer) / nrow(springer_nat) * 100`% of the overall Springer Nature article volume in the Open APC data, and `r sum(springer$euro) / sum(springer_nat$euro) * 100` % in terms of the amount being spent. Median publication fee spending slightly differed between Springer Science + Business Media (`r median(springer$euro)` €) and Nature Publishing (`r median(nature$euro)` €). However, price variation was higher for Nature Publishing journals (SD = `r sd(nature$euro)`€) compared with that of the former Springer Science + Business Media titles (SD = `r sd(springer$euro)` €).

In contrast to Springer Nature, other well-established publishing houses such as Elsevier and Wiley-Blackwell rank lower in our analysis. Table 2, furthermore, illustrates the price variations across and within publishers, which confirms earlier findings [@Solomon_2012; @Pinfield_2015].

@@

Prices also varied within single journals. Based on the number of articles paid for, Table 3 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals. Payments to these journals represent `r apc_journal %>% filter(!journal_full_title == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the overall expenditure. The multidisciplinary journal PLOS ONE and the journals New Journal of Physics, Atmospheric Chemistry and Physics Discussions and Frontiers in Psychology, all of which publish contributions from all branches of their respective discipline, were also well represented in the Open APC data-set. In general, an estimated 14 out of more than 10,000 journals registered in DOAJ in 2015 accounted for up to 15–20 % of all articles published in fully open access journals [@Bj_rk_2015] In the case of Atmospheric Chemistry and Physics Discussions, the large price range can be explained by the fact that this journal charges per page and also takes the submission's file format into consideration.

@@

In Germany, institutional spending on publication fees charged by open access journals has increased over the years. These findings are consistent with the general trend of publication fees as revenue source for open access publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. They also reflect the increasing importance of providing central support to cover these charges [@Fournier_2013; @Kingsley_2014]. Similar to investigations of the expenditure on publication fees at an institutional level in the UK [@Pinfield_2015], spending size varies across the German universities and research organisations included in the Open APC initiative. With a proportion of 39 % of the overall number of articles, the Max Planck Society, a large non-university research organisation, supported most open access journal articles. This likely reflects the centralised library support at the Max Planck Society where the Max Planck Digital Library has managed large open access agreements with publishers over the last decade [@Schimmer_2013, @Sikora_2015] leading not only to a large number of supported articles but also to an advantage in monitoring and thus being able to report on this data. Many universities and research organisations, on the other hand, reported a remarkably lower number of supported articles.

To answer data curation issues while unifying spending data raised by UK initiatives, re-using DOIs to gather bibliographic metadata from Crossref is a promising approach. In our study, Crossref thoroughly indexed open access journal articles disclosed in the Open APC data-set, providing publisher and journal titles for 99 % of the overall article volume. Making use of metadata from Crossref, therefore, reduces the extensive validation work of bibliographic information provided that the DOI is made available in the cost data, a requirement needed to be addressed also in other fields of the quantitative study of scholarly communication [@Haustein_2016]. Drawing on Crossref would also increase the comparability of cost data to prepare future negotiations with publishers on open access agreements, because Crossref's metadata reflects the current landscape of academic publishing in terms of ongoing mergers of publishing houses or name changes. Comparative studies of publication fee spending using data on the article-level would also benefit from such an approach.

This study is limited in some respects. One is that we cannot assess whether publishers and journals granted publication fee discounts because the Open APC initiative, dedicated to make self-reporting easy by using a minimal data schema, does not track this kind of information. However, large price ranges of particular journals suggests that varying pricing schemes are in place, an observation also made earlier [@Solomon_2012, @Pinfield_2015]. Adding to this complexity, it is likely that some institutions only paid parts of the publication fee. Take for instance the journal Nature Communication. Charges reported to the Open APC initiative ranged between 2000 €, the DFG price cap, and 4.403 €. Although such payments made from several budgets are a proposed strategy to sustain publication funds at German universities [@Fournier_2013], these pro rata amounts cannot be identified as such within the Open APC data, leading to a possibly flawed representations of publication fee spending in Germany. In another case, one university included its contributions to the SCOAP^3 consortia and presumably divided the sum by the articles published by their authors in SCOAP^3-covered journals.^[26] This is very arbitrary, since the mean APC for an institution can only be determined after the end of a full 3-year funding cycle.

It must also be noted that reporting to the Open APC initiative is voluntary. Therefore, not all institutions in Germany that provide central funding of publication fees contribute cost data to this initiative. In a qualitative survey, asking why German institutions are reluctant to share their cost data through the Open APC initiative, one institution feared that increase in transparency would allow publishers to adjust prices in their favour. Others pointed out that the workload to produce such a data-set could be too extensive [@deppe_2015]. As no complete registry of institutional open access funds or similar support structures exists, we cannot assess the number of non-participants in Germany.

>>>>

The study of institutional spending on open access journal articles has been limited for several reasons. One is that payment on these charges is fragmented across the budgets of grant agencies, research institutions, and libraries, or is covered by personal budgets. Asking 9,645 authors from various disciplines how they financed publication fees, a comprehensive survey in 2010 revealed that the majority of the respondents had access to research funding or institutional support to cover these charges. By contrast, 12 % paid publication fees individually [@soap]. These findings are consistent with that of other studies, adding that sources of funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets, on the other hand, are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

@@

Hybrid journals substantially add to this complexity of open access funding [@Bj_rk_learned_2014; @Kingsley_2014; @Pinfield_2015]. These journals, allowing articles to be published immediately as open access after a charge was paid, rely both on subscriptions and publication fees as revenue sources. Although the uptake of open access through hybrid journals was described as lower and more expensive compared to that of fully open access journals [@Solomon_2012; @Bj_rk_learned_2014], this model has gained increasing attention because of recent open access science policies, notably from the UK [@Pinfield_Review_2015].

To address these problems of fragmented spending on publication fees and of in-transparency about what was being paid, some European research funders and research performing institutions have recently begun to disclose their payment for publication as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc_12] and the Austrian Science Fund FWF [@fwf_apc_13]. The not-for-profit company Jisc followed by collecting data from UK universities [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report expenditure along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates and persistent identifiers to the full text including the Digital Object Identifier (DOI) [@Neylon_2014; @woodward_2014]. Parts of Jisc's cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variances, as well as a varying number of articles UK universities supported between 2007 - 2014, being consistent with earlier studies that collected price information from journal websites [@Solomon_2012].

@@

This paper focuses on how much German universities and research organisations spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for these charges. Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" program in 2011, only a few central funds existed [@eppelin_2012]. This is similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012]. The DFG has enforced a set of criteria grantees have to comply with  [@Fournier_2013], leading to related implementations for open access supporting across German universities. These criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 €[^3]. Grantees agree not only to pay for APCs, but also to find ways to improve the handling of those financial transactions. This includes central invoicing schemes and memberships, among others, agreed between university libraries and publishers that often lead to a discount on publication fees [@Fournier_2013].

Non-university research organisations, mainly those institutes organised in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft, are not eligible for this DFG funding program. But in response, some organisations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library (MPDL) [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The evolving institutional support structures to cover open access publication fees has led to calls for an unified approach towards supporting open access journal publishing in Germany. The MPDL called for re-allocating subscriptions in favour of financing open access charges in 2014 [@Schimmer_Geschuhn_Vogler]. The Allianz der Wissenschaftsorganisationen[^10c], a science policy board representing all major research organisations in Germany, marked price transparency as one way to sustain an "adequate open access publication system" [@allianz]. Reflecting Austrian and UK initiatives to share institutional spending on open access publication fees as open data, as well as professional discussions on open access publishing, Bielefeld University Library began to openly share its payment of publication fees in May 2014. After engaging with the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11], other German institutions joined under the umbrella of the Open APC initiative soon after [@openapc].

@@

The aim of the study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Using self-reported cost data from the Open APC initiative, the analysis focused on the amount that was being spent on publication fees, and compared these expenditure with data from related Austrian and UK initiatives, both in terms of size and the proportion of articles being published in fully and hybrid open access journals. We also asked how thoroughly self-reported articles were indexed in Crossref, a DOI minting agency for scholarly literature, and analysed how the institutional spending was distributed over publishers and journal titles.

@@

We analysed self-reported cost data released by the Open APC initiative on May 13, 2016,[^24] to assess institutional spending on open access publication fees in Germany. In addition to administrative data about the amount paid per article including value added tax, the reporting institution, and the year of invoicing, we used information about whether an article was published in a fully open access journal or in a hybrid journal as well as the DOI reported in the data-set.

We fetched bibliographic metadata for each article from Crossref on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from Crossref as well, this information was retrieved at the time when the participating institutions submitted their expenditure. The Open APC initiative kept track of the date when these data-sets were submitted with Git, a version control system, increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub to be transparent over time. However, during these data collection activities Crossref had regularly updated metadata to represent ongoing mergers of publishing houses. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve  updated bibliographic metadata from Crossref, and to merge these records with the administrative information instead of re-using the historical publisher and journal information contained in the Open APC data-set.

@@

Data collection also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by Germany universities and research organisations with that of other initiatives, we consulted the openly available data-sets from the the Austrian Science Fund FWF [@fwf_apc_14; @fwf_apc_15], Jisc [@jisc_14; @jisc_15] and the Wellcome Trust [@wellcome_apc_13_14; @wellcome_apc_14_15]. For analysis, we obtained the overall publication fee spending on both fully and hybrid open access journal articles. In the case of FWF, we gathered the cost information from the accompanying spending reports. We used the spreadsheet data to calculate Wellcome Trust's and Jisc's spending, and converted the prices from GBP to Euro in accordance with the average foreign exchange reference rates provided by the European Central Bank. Our comparison of the open data initiatives focussed on the last two years 2014 and 2015. Because Wellcome Trust's spending was reported for the fiscal periods 2013 - 2014 and 2014 - 2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded articles from the analysis for which  information about the cost or the journal type was missing. In the case of Jisc's 2014 data [@jisc_14], for instance, we excluded spending on 2,812 publications that amounted to 4.861.772 € from the analysis because no publication type was given in the data-set.

Data curation activities of the Open APC initiative and those of the other initiatives differed in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institutions were required to report, in case of the Wellcome Trust spending data, DOIs were also being identified by automated compliance checks. Our first screening of the data-sets revealed that some articles published in Crossref-indexed journals lacked a DOI. Because of these different methods to curate the cost data, and as our main focus was institutional funding for publication fees in Germany, we decided to only compare German APC spending with that reported by other initiatives. We, therefore, did not analyse the distribution of spending over publishers and journal titles, or the indexing coverage in Crossref for the Austrian and UK spending data.

@@

After excluding payments being made for non-journal articles as well as articles invoiced in 2016, we retrieved `r nrow(my_apc)` open access journal articles `r length(unique(my_apc$institution))` German universities and research institutions financially supported between 2005 and 2015. As illustrated in Figure 1, payments made for open access journal articles increased over the years. While one institution supported `r filter(my_apc, period == min(my_apc$period)) %>% nrow()` articles in `r as.character(min(my_apc$period))`, most institutions included in our study shared their expenditure from 2013 onwards. With `r max(apc_time$articles)` articles, the year `r apc_time[apc_time$articles == max(apc_time$articles),"period"]` was best represented. However, at the time of analysis, not all, but `r my_apc %>% filter(period == 2015) %>% group_by(institution) %>% summarise(n()) %>% nrow()` institutions contributed cost information, suggesting a time lag between payments made and reporting these spending to the Open APC initiative.

@@

Figure 3 presents institutional spending per article and year. Large price variations could be observed. Publication fees that were paid by the German universities and research organisations ranged between `r min(my_apc$euro)` € and `r max(my_apc$euro)` €  (SD `r sd(my_apc$euro)`). However, the average price paid varied somewhat during the period 2011 and 2015 (`r my_apc %>% filter(period == 2011) %>% summarise(mean(euro)) %>% round()` - `r my_apc %>% filter(period == 2015) %>% summarise(mean(euro)) %>% round()` €).

@@

The number of APC payments per institutions varied considerably (see Table 1). With `r filter(my_apc, institution == "MPG") %>% nrow()` reported articles, the Max Planck Society contributed `r (filter(my_apc, institution == "MPG") %>% nrow() / nrow(my_apc)) * 100` % of the overall article volume. By contrast, for several universities that only recently begun to set up centrally managed open access funds to cover publication fees we observed a lower number of support for open access journal articles.

@@

Table 2 compares Open APC spending data with that of the Austrian FWF, as well as with Jisc's and Wellcome Trust's expenditure. Prices were converted according to the average Euro exchange rate of the examined periods, and gathered for both fully and hybrid open access journals. The comparison revealed that the Open APC initiative lacked cost information about hybrid journals, whereas the related Austrian and UK open data initiatives reported a large share of spending on these journals between 2014 and 2015. Over the years 2005 – 2015, `r my_apc %>% group_by(institution) %>% filter(is_hybrid == TRUE) %>% summarise() %>% nrow()` out of `r length(unique(my_apc$institution))` German universities and research institutions reported `r my_apc %>% filter(is_hybrid == TRUE) %>% nrow()` hybrid journal articles to the Open APC initiative, representing `r (my_apc %>% filter(is_hybrid == TRUE) %>% nrow() / my_apc %>% nrow()) * 100` % of the overall supported article volume. In terms of the number of supported articles and the amount being spent on publication fees, by contrast, the Open APC data-set provided the most comprehensive price information for fully open access journals compared to what the Austrian and UK initiatives had reported.

@@

Comparison of average prices suggests that publishing in hybrid journal was more expensive than in fully open access journals. Price differences between these two categories were also reported earlier, indicating that prices for fully open access journals were on average lower [@Solomon_2012; @Pinfield_2015]. In 2014 and 2015, the mean price for fully open access journals calculated from all data-sets was below the DFG price cap of 2,000 €.

@@

We used the DOI to automatically fetch publisher and journal names for each article from the Crossref REST API. Table 2 shows the top ten publishers in terms of payments made. They represented `r apc_publisher %>% filter(!publisher == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the overall spending on publication fees. In total, payments were made to `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. In comparison with data from the UK, a greater share on total spending of some open access publishers could be observed. @Pinfield_2015, for instance, reported remarkably lower numbers for the open access publishers MPDI, Copernicus, and Hindawi.

@@

Most of publication fee spending in Germany was on articles published in Springer Nature journals, likely profiting from the merge with the open access publisher BioMed Central in 2008, and that between the well-established publishers Springer Science + Business Media and Nature Publishing Group in 2015. Using the Crossref-Member-ID instead of the publisher name, we were able to differentiate between journals formely published by Springer Science + Business Media and Nature Publishing Group. In terms of articles, the majority of payments made were for publications in journals formerly associated with Springer Science + Business Media. Springer Science + Business Media journals accounted for `r nrow(springer)` articles, representing `r nrow(springer) / nrow(springer_nat) * 100`% of the overall Springer Nature article volume recorded by the Open APC initiative, and `r sum(springer$euro) / sum(springer_nat$euro) * 100` % in terms of the amount being spent. Median publication fee spending slightly differed between Springer Science + Business Media (`r median(springer$euro)` €) and Nature Publishing (`r median(nature$euro)` €). However, price variation was higher for Nature Publishing journals (SD = `r sd(nature$euro)`€) compared with that of the former Springer Science + Business Media titles (SD = `r sd(springer$euro)` €).

In contrast to Springer Nature, other well-established publishing houses such as Elsevier and Wiley-Blackwell ranked lower in our analysis.

@@

Prices also varied within journals. Based on the number of articles paid for, Table 3 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals. We normalised PLOS journal titles, because the name chnages from "PLoS" to "PLOS" was only partly represented in the Crossref metadata at the time of our study. Payments to the top ten journals represent `r apc_journal %>% filter(!journal_full_title == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the overall expenditure. The multidisciplinary journal PLOS ONE and the journals New Journal of Physics, Atmospheric Chemistry and Physics Discussions and Frontiers in Psychology, all of which publish contributions from all branches of their respective discipline, were also well represented in the Open APC data-set. In general, an estimated 14 out of more than 10,000 journals registered in DOAJ in 2015 accounted for up to 15–20 % of all articles published in fully open access journals [@Bj_rk_2015] In the case of Atmospheric Chemistry and Physics Discussions, the large price range can be explained by the fact that this journal charges per page and also takes the submission's file format into consideration.

@@

In Germany, institutional spending on publication fees charged by open access journals has increased over the years. These findings are consistent with the general trend of publication fees as revenue source for open access publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. They also reflect the evolving institutional support to cover these charges in Germany [@Fournier_2013]. Similar to investigations of the expenditure on publication fees at an institutional level in the UK [@Pinfield_2015], spending size varies across the German universities and research organisations included in the Open APC initiative. With a proportion of 39 % of the overall article volume, the Max Planck Society, a large non-university research organisation, supported most open access journal publications. This likely reflects the centralised library support at the Max Planck Society where the Max Planck Digital Library has managed large open access agreements with publishers over the last decade [@Schimmer_2013, @Sikora_2015]. This presumably not only resulted in a large number of supported articles, but also contributed to an advantage in monitoring and thus being able to report on this data. Many universities and research organisations, on the other hand, reported a remarkably lower number of supported articles.

To answer data curation issues while unifying UK spending data [@Neylon_2014; @woodward_2014], re-using DOIs to gather bibliographic metadata from Crossref is a promising approach. In our study, Crossref thoroughly indexed open access journal articles disclosed in the Open APC data-set, providing publisher and journal titles for 99 % of the overall article volume. Making use of metadata from Crossref, therefore, reduces the extensive validation work of bibliographic information provided that the DOI is made available in the cost data. DOI availability is a requirement needed to be addressed in other fields of the quantitative study of scholarly communication as well, such as in altmetrics to deal with heterogeneous data [@Haustein_2016]. Drawing on Crossref has the potential to increase the comparability of cost data to prepare future negotiations with publishers on open access agreements, because Crossref's metadata reflects the current landscape of academic publishing in terms of ongoing mergers of publishing houses. Future comparative studies of publication fee spending using data on the article-level also benefit from such an approach.

This study is limited in some respects. One is that we cannot assess whether publishers and journals granted publication fee discounts because the Open APC initiative uses a minimal data scheme to make self-reporting easy, and, therefore, does not track this kind of information. However, large price ranges of particular journals suggests that varying pricing schemes are in place, an observation also made earlier [@Solomon_2012, @Pinfield_2015]. Adding to this complexity, it is likely that some institutions only paid parts of the publication fee. Take for instance the journal Nature Communication. Charges reported to the Open APC initiative ranged between 2000 €, the DFG price cap, and 4.403 €. Although such payments made from several budgets are a proposed strategy to sustain publication funds at German universities [@Fournier_2013], these pro rata amounts cannot be identified as such within the Open APC data, leading to a possibly flawed representations of publication fee spending in Germany. In another case, one university included its contributions to the SCOAP^3 consortia and presumably divided the sum by the articles published by their authors in SCOAP^3-covered journals.^[26] This is very arbitrary, since the mean APC for an institution can only be determined after the end of a full 3-year funding cycle. Other factors affecting price variations are currency movements over the years, and different tax rates in Germany. For instance, the Max Planck Society has a limited input tax reduction. The refund of input value added tax for publication fees is 20%. Local and payment currency, as well as taxes, as suggested by @Pinfield_2015 to increase the transparency of publication fee spending, were not disclosed in the Open APC initiative's data-set at the time of our study.

It must also be noted that reporting to the Open APC initiative is voluntary. Therefore, not all institutions in Germany that provide central funding of publication fees contribute cost data to this initiative. In a qualitative survey, asking why German institutions are reluctant to share their cost data through the Open APC initiative, one institution feared that an increase in transparency would allow publishers to adjust prices in their favour. Others pointed out that the workload to produce such a data-set could be too extensive [@deppe_2015]. As no reliable registry of institutional open access funds or similar support structures in Germany exists, we cannot assess the number of non-participants in Germany.

Our analysis of how institutional spending per articles was distributed over publishers and journal titles indicates that open access publishing is heterogeneous and concentrated at the same time. While we were able to identify 139 individual publishing houses that were supported by the German universities and research organizations, the distribution is highly skewed. Ten publishers collected 92 % of open access publication fee spending, being consistent with an observed high concentration of few publishers in current academic publishing [@Larivi_re_2015]. However, our study could not confirm that publications in open access journals owned by traditional publishing houses account for most of the spending on publication fees as observed by @Pinfield_2015. Rather, open access publishers such as Public Library of Science (PloS), Copernicus or MPDI rank higher in our study than in the analyses of cost data in the UK.
